canaccessR: An open data product for analyzing transportation accessibility to employment and grocery stores in Canada's largest metropolitan areas.

### 2025-03-07 paper/Manuscript-Data-Package.Rmd
Data and methods
Finished data sources and methods; added other ODP references; descriptive statistics and how to use still incomplete

---
title: "canaccessR: An open data product for analyzing transportation accessibility to employment and grocery stores in Canada’s largest metropolitan areas."
author:
  - name: João Pedro Figueira Amorim Parga
    email: joao.parga@mail.utoronto.ca
    affil-id: 1
    orcid: 0000-0002-4105-5927
    footnote: 1
  - name: Anastasia Soukhov
    email: soukhoa@mcmaster.ca
    affil-id: 2
    orcid: 0000-0003-4371-4831
  - name: Robert Nutifafa Arku
    email: rn.arku@mail.utoronto.ca
    affil-id: 1
    orcid: 0000-0002-2018-886X
  - name: Christopher Higgins
    email: cd.higgins@utoronto.ca
    affil-id: 3
    orcid: 0000-0002-3551-7750
  - name: Antonio Páez
    email: paezha@mcmaster.ca
    affil-id: 2
    orcid: 0000-0001-6912-9919
affiliations:
  - id: 1
    name: University of Toronto
  - id: 2
    name: McMaster University
  - id: 3
    name: University of Toronto Scarborough
# address:
#   - code: CODE
#     address: "Some school"
footnote:
  - code: 1
    text: "Corresponding Author"
abstract: 
  "In this paper, we describe the {canaccesR} package, an open data product (ODPs) created using the R statistical language. {canaccess} is a data package that provides public transit travel time estimates (travel time matrices - TTM) to employment locations and grocery stores across the 12 largest Canadian metropolitan areas. We calculate these estimates for each Dissemination Area (DA) within these regions for the years 2019 and 2023. To do so, we use the {r5r} R package, General Transit Feed Specification (GTFS), OpenStreetMap (OSM), DMTI’s Enhanced Points of Interest, and Statistics Canada Census data. This data package can be used by researchers, practitioners, and transit agencies to estimate accessibility to essential services across these regions. These estimates can be used to compare different regions across Canada in terms of their accessibility and to conduct within-region equity assessments regarding access to services, which can inform improvements in transportation policies related to accessibility. The package is still in its initial phase and may undergo expansions in the future by adding TTM's for other destinations (e.g., schools, healthcare facilities). Finally, as an ODPs, the {canaccess} package allows for open exploration, use, and contribution by users through its GitHub repository."

@@

The objective of this paper is to describe the {canaccesR} open data package.
Its main contents are a set of public transit travel time matrices (TTM) estimates to employment and groceries stores from the 12 largest Canadian metropolitan areas in the years 2019 and 2023.
These estimates were created by leveraging expertise in data science, computer programming, and transportation accessibility, using the {r5r} R package [@pereiraR5rRapidRealistic2021].
We used transport and street network, population, and employment data from different sources to estimate these TTM's datasets.

Estimating accessibility - the potential offered by the transportation system to reach destinations [@paezMeasuringAccessibilityPositive2012] - requires specialized datasets and technical expertise.
Despite recent advancements in the field, ready-to-use available data on transportation accessibility is still sparse.
Thus, we leverage technical knowledge to make publicly available ready-to-use data on transportation accessibility.
To create the package, we integrated and processed raw data from diverse sources, estimated TTM's for two destinations types (e.g., jobs and groceries) across the largest cities in the country, and distributed these findings through this transparent and open source data product.
Our main contribution is to provide analysis-ready data for Canada's largest cities, thus making the fields of accessibility research and urban analytics more accessible.

The package's main audiences are Canadian researchers in urban planning and transportation and transportation system agencies.
We anticipate three main uses for the open data product (ODP) described in this paper.
First, the datasets allow for static assessment of the level of public transit accessibility across the country's largest cities before and after the COVID-19 pandemic.
In other words, we make it easier for those interested to compare cities and their level of public transit accessibility to essential destinations (such as employment centers and groceries stores).
Second, researchers can use the package's information as inputs to evaluate changes in accessibility throughout this period. 
Researchers may also investigate disparities in these changes across space within those cities, given the spacial character of the datasets made available here.
Third, as is common practice in transportation accessibility research, these evaluations can substantiate broader investigations on matters of justice and equity in transportation.
For example, the TTM estimates allow for evaluating the evolution of public transit's accessibility by income or spatial distribution across all Dissemination Areas (DA's) of each of the 12 cities in the sample [@pargaDemocraticAccessOur2024].

@@

# Data Sources

The locations included in the data package comprise the 12 largest (population-wise) Census metropolitan areas (CMA's) based on the 2021 Canadian Census [@governmentofcanada2021CensusPopulation2021] [^1]. 
These locations are Toronto, Montreal, Vancouver, Ottawa-Gatineau, Calgary, Edmonton, Quebec City, Winnipeg, Hamilton, Kitchener-Cambridge-Waterloo, London, and Halifax.

[^1]: We included Oshawa, Ontario, as part of the Greater Toronto Area (GTA) because of the former's proximity to the latter. We also included Abbotsford-Mission, British Columbia, as part of the Vancouver metropolitan area because of the former's proximity to a transit station on the region’s West Coast Express commuter rail line.
We used four main data sources to construct the {canaccesR} data package: General Transit Feed Specification (GTFS), OpenStreetMap (OSM), DMTI’s Enhanced Points of Interest, and Statistics Canada Census data.

<!-- ## General Transit Feed Specification -->
We used the GTFS files from all transit agencies within the selected CMA's to obtain the information on their public transit schedule from 2019 to 2023.
<!-- ## OpenStreetMaps -->
The OpenStreetMap data provided information on the transit network and was necessary for the routing used to estimate the travel time matrices.
<!-- ## Census Data -->
We used the 2016 and 2021 Canadian Census' information on the spatial distribution of the population and the number of workplace locations (employment) across the CMA's [@governmentofcanada2016CensusPopulation2016, governmentofcanada2021CensusPopulation2021].
We aggregayed all the resulting travel time matrices at the Dissemination Area (DA) level, which comprise the fundamental unit of analysis in data package.
<!-- ## DMTI’s Enhanced Points of Interest -->
Finally, we use the DMTI’s Enhanced Points of Interest dataset to obtain the location of the groceries stores within every urban area selected [@dmtispatialincEnhancedPointsInterest2015]. 
We filtered the locations within the DMTI dataset using the grocery stores code from the North American Industry Classification System (NAICS) and the Standard Industrial Classification (SIC).

# Data processing

Routing, cleaning, etc.

@@

In this paper, we describe the {canaccesR} data package, created using the {r5r} R package and transit schedule, street network, employment, and population data.
The package's main contents refers to the ready-to-use travel time matrices for public transit to reach employment and groceries stores in Canada's 12 largest urban areas.
We expect the contents of the package to be used in transportation accessibility evaluations within and across those regions.
Moreover, these datasets can be used in further equity assessments that evaluate the distribution of accessibility across space and between social groups.
In other words, we hope that by making these datasets publicly available, future analysis can contribute to making Canada's transportation system more just and fair, considering accessibility's as the main social good of transportation [@martensTransportJusticeDesigning2016], and the inherent connection between public transit and the "right to the city" [@cogginRightTransportMoving2015].

>>>>

---
title: "canaccessR: An open data product for analyzing transportation accessibility to employment and grocery stores in Canada’s largest metropolitan areas."
author:
  - name: João Pedro Figueira Amorim Parga
    email: joao.parga@mail.utoronto.ca
    affil-id: 1
    orcid: 0000-0002-4105-5927
    footnote: 1
  - name: Anastasia Soukhov
    email: soukhoa@mcmaster.ca
    affil-id: 2
    orcid: 0000-0003-4371-4831
  - name: Robert Nutifafa Arku
    email: rn.arku@mail.utoronto.ca
    affil-id: 1
    orcid: 0000-0002-2018-886X
  - name: Christopher Higgins
    email: cd.higgins@utoronto.ca
    affil-id: 3
    orcid: 0000-0002-3551-7750
  - name: Antonio Páez
    email: paezha@mcmaster.ca
    affil-id: 2
    orcid: 0000-0001-6912-9919
affiliations:
  - id: 1
    name: University of Toronto
  - id: 2
    name: McMaster University
  - id: 3
    name: University of Toronto Scarborough
# address:
#   - code: CODE
#     address: "Some school"
footnote:
  - code: 1
    text: "Corresponding Author"
abstract: 
  "In this paper, we describe the {canaccesR} package, an open data product (ODPs) created using the R statistical language. {canaccess} is a data package that disseminates public transit travel time estimates (travel time matrices - TTM) to employment locations and grocery stores across the 12 largest Canadian metropolitan areas. We calculate these estimates for each Dissemination Area (DA) within these regions for the years 2019 and 2023. To do so, we use the {r5r} R package, General Transit Feed Specification (GTFS), OpenStreetMap (OSM), DMTI’s Enhanced Points of Interest, and Statistics Canada Census data. This data package can be used by researchers, practitioners, and transit agencies to estimate accessibility to essential services across these regions. These estimates can be used to compare different regions across Canada in terms of their accessibility and to conduct within-region equity assessments regarding access to services, which can inform improvements in transportation policies related to accessibility. The package is still in its initial phase and may undergo expansions in the future by adding TTM's for other destinations (e.g., schools, healthcare facilities). Finally, as an ODPs, the {canaccess} package allows for open exploration, use, and contribution by users through its GitHub repository."

@@

The objective of this paper is to describe the {canaccesR} open data package.
Its main contents are a set of public transit travel time matrices (TTM) estimates to employment and groceries stores from the 12 largest Canadian metropolitan areas in 2019 and 2023.
These estimates were created by leveraging expertise in data science, computer programming, and transportation accessibility using the {r5r} R package [@pereiraR5rRapidRealistic2021].
We used public transit schedule, transport network, population, and business location data from different sources to estimate these TTM's datasets.

Estimating accessibility - the potential offered by the transportation system to reach destinations [@paezMeasuringAccessibilityPositive2012] - requires specialized datasets and technical expertise.
Recent efforts following an open-source and transparent philosophy have been made to disseminate useful data and information on transportation in the Canadian context [@soukhovTTS2016RDataSet2023]. 
However, despite these initiatives, pre-processed and available data on transportation accessibility is scarse.
Within this context, we expect to help filling this gap by the processing raw data into user-ready data and making them publicly available to advance knowledge on the field.

To create the package, we integrated and processed raw data from diverse sources, estimated TTM's for two destinations types (*e.g.*, jobs and groceries) across the largest cities in the country, and distributed these findings through this transparent and open source data product.
Our main contribution is to provide analysis-ready data for Canada's largest cities on the topic of transportation accessibility, thus making urban analytics in the country more accessible and contributing to future research.

The package's main audiences are Canadian researchers in urban planning and transportation and transportation system agencies.
We anticipate three primary uses for the open data product (ODP) described in this paper.
First, the datasets allow for static assessment of the level of public transit accessibility across the country's largest cities before and after the COVID-19 pandemic.
In other words, {canaccesR} makes it easier for those interested in comparing cities regarding their level of public transit accessibility to essential destinations (such as employment centers and groceries stores) to do so.
Second, the temporal and spatial characters of the datasets made available here allow researchers to evaluate accessibility changes through time and across space within the largest Canadian urban areas.
Third, as is now common practice in transportation accessibility research, used as inputs, these estimates can substantiate broader investigations on transportation justice and equity [@higginsChangesAccessibilityEmergency2021; @humbertoHowTranslateJustice2023; @pereiraGeographicAccessCOVID192021].
For example, the TTM estimates allow for evaluating the evolution of public transit's accessibility by income or spatial distribution across all Dissemination Areas (DA's) of each of the 12 cities in the sample [@pargaDemocraticAccessOur2024].
In other words, the package's contents can be used from straightforward assessments of accessibility in Canadian urban areas to more theoretically and morally complex evaluations of justice in the country's urban transportation system.

@@

# Data and methods

## Raw data sources

The locations included in the data package comprise the 12 largest (population-wise) Census metropolitan areas (CMA's) based on the 2021 Canadian Census [@governmentofcanada2021CensusPopulation2021] [^1]. 
These locations are Toronto, Montreal, Vancouver, Ottawa-Gatineau, Calgary, Edmonton, Quebec City, Winnipeg, Hamilton, Kitchener-Cambridge-Waterloo, London, and Halifax.
We used four main data sources to construct the {canaccesR} data package: General Transit Feed Specification (GTFS), OpenStreetMap (OSM), DMTI’s Enhanced Points of Interest, and Statistics Canada Census data.

[^1]: We included Oshawa, Ontario, as part of the Greater Toronto Area (GTA) because of the former's proximity to the latter. We also included Abbotsford-Mission, British Columbia, as part of the Vancouver metropolitan area because of the former's proximity to a transit station on the region’s West Coast Express commuter rail line.

<!-- ## General Transit Feed Specification -->
We manually collected and processed the GTFS files from all transit agencies within the selected CMA's to use their information on the public transit schedule in 2019 and 2023.
<!-- ## OpenStreetMaps -->
The OpenStreetMap data for the selected areas were collected through the {osmextract} package [@gilardiOsmextractDownloadImport2025].
We used OSM data from 2019 and 2023, which provided information on the areas' transit network in two points in time.
<!-- ## Census Data -->
We collected data from the 2016 Canadian Census using the {cancensus} R package [@vonbergmannCancensusPackageAccess2022] and used its information on the spatial distribution of the population and the number of workplace locations (employment) across the CMA's [@governmentofcanada2016CensusPopulation2016].
<!-- ## DMTI’s Enhanced Points of Interest -->
Finally, we gathered and cleaned the 2023 DMTI’s Enhanced Points of Interest dataset to obtain the location of the groceries stores within every urban area selected [@dmtispatialincEnhancedPointsInterest2015]. 
We filtered the locations within the DMTI dataset using the grocery stores code from the North American Industry Classification System (NAICS) and the Standard Industrial Classification (SIC).
{**THE CODE FOR THIS ESTIMATIONS AND THE TRAVEL TIME MATRICES IS ON THE transit_death_spiral github repo. 1) SHOULD WE CITE IT? 2) IS THAT A PROBLEM?**}

## Methods: travel time matrices processing

Using the {r5r} R package, we estimated public transit travel times for two destination types, grocery stores and jobs. 
For each amenity type, we chose a likely travel time and day of the week. 
We set a 15 minutes time window and the maximum trip duration to 120 minutes. 
For groceries stores, we set the departure date to a weekend afternoon and the departure time to between 12:00 PM to 12:15 PM on April 20, 2019 and April 22, 2023.
For employment, we ran the analysis on a typical weekday morning rush-hour commute, more specifically 8:00 to 8:15 AM departure on Tuesday, April 16, 2019 and Tuesday, April 18, 2023 [^3].
In both cases, we assumed that walking was the mode of travel from origin to transit stop and from transit stop to destination.
We aggregated all the resulting travel time matrices at the Dissemination Area (DA) level, which comprise the fundamental unit of analysis in data package.
[^3]: The one exception is Quebec City, where the routing for 2019 occurs on a Saturday and Tuesday in June (instead of April) due to the GTFS data unavailability.

@@

In this paper, we describe the {canaccesR} data package, created using the {r5r} R package and transit schedule, street network, employment, and population data.
The package's main contents refers to the ready-to-use travel time matrices for public transit to reach employment and groceries stores in Canada's 12 largest urban areas.
We expect the contents of the package to be used in transportation accessibility evaluations within and across those regions.
Moreover, these datasets can be used in further equity assessments that evaluate the distribution of accessibility across space and between social groups.
Furthermore, in the spirit of open data products [@arribas-belOpenDataProductsA2021], the package can be expanded through collaboration with other researchers by, for example, including travel time matrices to other essential destinations within the DMTI's dataset (*e.g.*, schools, healthcare, etc.).
In other words, we hope that by making these datasets publicly available, future analysis can contribute to making Canada's transportation system more just and fair, considering accessibility's as the main social good of transportation [@martensTransportJusticeDesigning2016], and the inherent connection between public transit and the "right to the city" [@cogginRightTransportMoving2015].

### 2025-03-07 paper/Manuscript-Data-Package.tex
Data and methods
Finished data sources and methods; added other ODP references; descriptive statistics and how to use still incomplete

\begin{abstract}
In this paper, we describe the \{canaccesR\} package, an open data
product (ODPs) created using the R statistical language. \{canaccess\}
is a data package that provides public transit travel time estimates
(travel time matrices - TTM) to employment locations and grocery stores
across the 12 largest Canadian metropolitan areas. We calculate these
estimates for each Dissemination Area (DA) within these regions for the
years 2019 and 2023. To do so, we use the \{r5r\} R package, General
Transit Feed Specification (GTFS), OpenStreetMap (OSM), DMTI's Enhanced
Points of Interest, and Statistics Canada Census data. This data package
can be used by researchers, practitioners, and transit agencies to
estimate accessibility to essential services across these regions. These
estimates can be used to compare different regions across Canada in
terms of their accessibility and to conduct within-region equity
assessments regarding access to services, which can inform improvements
in transportation policies related to accessibility. The package is
still in its initial phase and may undergo expansions in the future by
adding TTM's for other destinations (e.g., schools, healthcare
facilities). Finally, as an ODPs, the \{canaccess\} package allows for
open exploration, use, and contribution by users through its GitHub
repository.
\end{abstract}

@@

The objective of this paper is to describe the \{canaccesR\} open data
package. Its main contents are a set of public transit travel time
matrices (TTM) estimates to employment and groceries stores from the 12
largest Canadian metropolitan areas in the years 2019 and 2023. These
estimates were created by leveraging expertise in data science, computer
programming, and transportation accessibility, using the \{r5r\} R
package \citep{pereiraR5rRapidRealistic2021}. We used transport and
street network, population, and employment data from different sources
to estimate these TTM's datasets.

Estimating accessibility - the potential offered by the transportation
system to reach destinations
\citep{paezMeasuringAccessibilityPositive2012} - requires specialized
datasets and technical expertise. Despite recent advancements in the
field, ready-to-use available data on transportation accessibility is
still sparse. Thus, we leverage technical knowledge to make publicly
available ready-to-use data on transportation accessibility. To create
the package, we integrated and processed raw data from diverse sources,
estimated TTM's for two destinations types (e.g., jobs and groceries)
across the largest cities in the country, and distributed these findings
through this transparent and open source data product. Our main
contribution is to provide analysis-ready data for Canada's largest
cities, thus making the fields of accessibility research and urban
analytics more accessible.

The package's main audiences are Canadian researchers in urban planning
and transportation and transportation system agencies. We anticipate
three main uses for the open data product (ODP) described in this paper.
First, the datasets allow for static assessment of the level of public
transit accessibility across the country's largest cities before and
after the COVID-19 pandemic. In other words, we make it easier for those
interested to compare cities and their level of public transit
accessibility to essential destinations (such as employment centers and
groceries stores). Second, researchers can use the package's information
as inputs to evaluate changes in accessibility throughout this period.
Researchers may also investigate disparities in these changes across
space within those cities, given the spacial character of the datasets
made available here. Third, as is common practice in transportation
accessibility research, these evaluations can substantiate broader
investigations on matters of justice and equity in transportation. For
example, the TTM estimates allow for evaluating the evolution of public
transit's accessibility by income or spatial distribution across all
Dissemination Areas (DA's) of each of the 12 cities in the sample
\citep{pargaDemocraticAccessOur2024}.

@@

\section{Data Sources}\label{data-sources}

The locations included in the data package comprise the 12 largest
(population-wise) Census metropolitan areas (CMA's) based on the 2021
Canadian Census \citep{governmentofcanada2021CensusPopulation2021}
\footnote{We included Oshawa, Ontario, as part of the Greater Toronto
  Area (GTA) because of the former's proximity to the latter. We also
  included Abbotsford-Mission, British Columbia, as part of the
  Vancouver metropolitan area because of the former's proximity to a
  transit station on the region's West Coast Express commuter rail line.
  We used four main data sources to construct the \{canaccesR\} data
  package: General Transit Feed Specification (GTFS), OpenStreetMap
  (OSM), DMTI's Enhanced Points of Interest, and Statistics Canada
  Census data.}. These locations are Toronto, Montreal, Vancouver,
Ottawa-Gatineau, Calgary, Edmonton, Quebec City, Winnipeg, Hamilton,
Kitchener-Cambridge-Waterloo, London, and Halifax.

We used the GTFS files from all transit agencies within the selected
CMA's to obtain the information on their public transit schedule from
2019 to 2023. The OpenStreetMap data provided information on the transit
network and was necessary for the routing used to estimate the travel
time matrices. We used the 2016 and 2021 Canadian Census' information on
the spatial distribution of the population and the number of workplace
locations (employment) across the CMA's
\citep[governmentofcanada2021CensusPopulation2021]{governmentofcanada2016CensusPopulation2016}.
We aggregayed all the resulting travel time matrices at the
Dissemination Area (DA) level, which comprise the fundamental unit of
analysis in data package. Finally, we use the DMTI's Enhanced Points of
Interest dataset to obtain the location of the groceries stores within
every urban area selected
\citep{dmtispatialincEnhancedPointsInterest2015}. We filtered the
locations within the DMTI dataset using the grocery stores code from the
North American Industry Classification System (NAICS) and the Standard
Industrial Classification (SIC).

\section{Data processing}\label{data-processing}

Routing, cleaning, etc.

@@

In this paper, we describe the \{canaccesR\} data package, created using
the \{r5r\} R package and transit schedule, street network, employment,
and population data. The package's main contents refers to the
ready-to-use travel time matrices for public transit to reach employment
and groceries stores in Canada's 12 largest urban areas. We expect the
contents of the package to be used in transportation accessibility
evaluations within and across those regions. Moreover, these datasets
can be used in further equity assessments that evaluate the distribution
of accessibility across space and between social groups. In other words,
we hope that by making these datasets publicly available, future
analysis can contribute to making Canada's transportation system more
just and fair, considering accessibility's as the main social good of
transportation \citep{martensTransportJusticeDesigning2016}, and the
inherent connection between public transit and the ``right to the city''
\citep{cogginRightTransportMoving2015}.

>>>>

\begin{abstract}
In this paper, we describe the \{canaccesR\} package, an open data
product (ODPs) created using the R statistical language. \{canaccess\}
is a data package that disseminates public transit travel time estimates
(travel time matrices - TTM) to employment locations and grocery stores
across the 12 largest Canadian metropolitan areas. We calculate these
estimates for each Dissemination Area (DA) within these regions for the
years 2019 and 2023. To do so, we use the \{r5r\} R package, General
Transit Feed Specification (GTFS), OpenStreetMap (OSM), DMTI's Enhanced
Points of Interest, and Statistics Canada Census data. This data package
can be used by researchers, practitioners, and transit agencies to
estimate accessibility to essential services across these regions. These
estimates can be used to compare different regions across Canada in
terms of their accessibility and to conduct within-region equity
assessments regarding access to services, which can inform improvements
in transportation policies related to accessibility. The package is
still in its initial phase and may undergo expansions in the future by
adding TTM's for other destinations (e.g., schools, healthcare
facilities). Finally, as an ODPs, the \{canaccess\} package allows for
open exploration, use, and contribution by users through its GitHub
repository.
\end{abstract}

@@

The objective of this paper is to describe the \{canaccesR\} open data
package. Its main contents are a set of public transit travel time
matrices (TTM) estimates to employment and groceries stores from the 12
largest Canadian metropolitan areas in 2019 and 2023. These estimates
were created by leveraging expertise in data science, computer
programming, and transportation accessibility using the \{r5r\} R
package \citep{pereiraR5rRapidRealistic2021}. We used public transit
schedule, transport network, population, and business location data from
different sources to estimate these TTM's datasets.

Estimating accessibility - the potential offered by the transportation
system to reach destinations
\citep{paezMeasuringAccessibilityPositive2012} - requires specialized
datasets and technical expertise. Recent efforts following an
open-source and transparent philosophy have been made to disseminate
useful data and information on transportation in the Canadian context
\citep{soukhovTTS2016RDataSet2023}. However, despite these initiatives,
pre-processed and available data on transportation accessibility is
scarse. Within this context, we expect to help filling this gap by the
processing raw data into user-ready data and making them publicly
available to advance knowledge on the field.

To create the package, we integrated and processed raw data from diverse
sources, estimated TTM's for two destinations types (\emph{e.g.}, jobs
and groceries) across the largest cities in the country, and distributed
these findings through this transparent and open source data product.
Our main contribution is to provide analysis-ready data for Canada's
largest cities on the topic of transportation accessibility, thus making
urban analytics in the country more accessible and contributing to
future research.

The package's main audiences are Canadian researchers in urban planning
and transportation and transportation system agencies. We anticipate
three primary uses for the open data product (ODP) described in this
paper. First, the datasets allow for static assessment of the level of
public transit accessibility across the country's largest cities before
and after the COVID-19 pandemic. In other words, \{canaccesR\} makes it
easier for those interested in comparing cities regarding their level of
public transit accessibility to essential destinations (such as
employment centers and groceries stores) to do so. Second, the temporal
and spatial characters of the datasets made available here allow
researchers to evaluate accessibility changes through time and across
space within the largest Canadian urban areas. Third, as is now common
practice in transportation accessibility research, used as inputs, these
estimates can substantiate broader investigations on transportation
justice and equity
\citep{higginsChangesAccessibilityEmergency2021, humbertoHowTranslateJustice2023, pereiraGeographicAccessCOVID192021}.
For example, the TTM estimates allow for evaluating the evolution of
public transit's accessibility by income or spatial distribution across
all Dissemination Areas (DA's) of each of the 12 cities in the sample
\citep{pargaDemocraticAccessOur2024}. In other words, the package's
contents can be used from straightforward assessments of accessibility
in Canadian urban areas to more theoretically and morally complex
evaluations of justice in the country's urban transportation system.

@@

\section{Data and methods}\label{data-and-methods}

\subsection{Raw data sources}\label{raw-data-sources}

The locations included in the data package comprise the 12 largest
(population-wise) Census metropolitan areas (CMA's) based on the 2021
Canadian Census \citep{governmentofcanada2021CensusPopulation2021}
\footnote{We included Oshawa, Ontario, as part of the Greater Toronto
  Area (GTA) because of the former's proximity to the latter. We also
  included Abbotsford-Mission, British Columbia, as part of the
  Vancouver metropolitan area because of the former's proximity to a
  transit station on the region's West Coast Express commuter rail line.}.
These locations are Toronto, Montreal, Vancouver, Ottawa-Gatineau,
Calgary, Edmonton, Quebec City, Winnipeg, Hamilton,
Kitchener-Cambridge-Waterloo, London, and Halifax. We used four main
data sources to construct the \{canaccesR\} data package: General
Transit Feed Specification (GTFS), OpenStreetMap (OSM), DMTI's Enhanced
Points of Interest, and Statistics Canada Census data.

We manually collected and processed the GTFS files from all transit
agencies within the selected CMA's to use their information on the
public transit schedule in 2019 and 2023. The OpenStreetMap data for the
selected areas were collected through the \{osmextract\} package
\citep{gilardiOsmextractDownloadImport2025}. We used OSM data from 2019
and 2023, which provided information on the areas' transit network in
two points in time. We collected data from the 2016 Canadian Census
using the \{cancensus\} R package
\citep{vonbergmannCancensusPackageAccess2022} and used its information
on the spatial distribution of the population and the number of
workplace locations (employment) across the CMA's
\citep{governmentofcanada2016CensusPopulation2016}. Finally, we gathered
and cleaned the 2023 DMTI's Enhanced Points of Interest dataset to
obtain the location of the groceries stores within every urban area
selected \citep{dmtispatialincEnhancedPointsInterest2015}. We filtered
the locations within the DMTI dataset using the grocery stores code from
the North American Industry Classification System (NAICS) and the
Standard Industrial Classification (SIC). \{\textbf{THE CODE FOR THIS
ESTIMATIONS AND THE TRAVEL TIME MATRICES IS ON THE
transit\_death\_spiral github repo. 1) SHOULD WE CITE IT? 2) IS THAT A
PROBLEM?}\}

\subsection{Methods: travel time matrices
processing}\label{methods-travel-time-matrices-processing}

Using the \{r5r\} R package, we estimated public transit travel times
for two destination types, grocery stores and jobs. For each amenity
type, we chose a likely travel time and day of the week. We set a 15
minutes time window and the maximum trip duration to 120 minutes. For
groceries stores, we set the departure date to a weekend afternoon and
the departure time to between 12:00 PM to 12:15 PM on April 20, 2019 and
April 22, 2023. For employment, we ran the analysis on a typical weekday
morning rush-hour commute, more specifically 8:00 to 8:15 AM departure
on Tuesday, April 16, 2019 and Tuesday, April 18, 2023 {[}\^{}3{]}. In
both cases, we assumed that walking was the mode of travel from origin
to transit stop and from transit stop to destination. We aggregated all
the resulting travel time matrices at the Dissemination Area (DA) level,
which comprise the fundamental unit of analysis in data package.
{[}\^{}3{]}: The one exception is Quebec City, where the routing for
2019 occurs on a Saturday and Tuesday in June (instead of April) due to
the GTFS data unavailability.

@@

In this paper, we describe the \{canaccesR\} data package, created using
the \{r5r\} R package and transit schedule, street network, employment,
and population data. The package's main contents refers to the
ready-to-use travel time matrices for public transit to reach employment
and groceries stores in Canada's 12 largest urban areas. We expect the
contents of the package to be used in transportation accessibility
evaluations within and across those regions. Moreover, these datasets
can be used in further equity assessments that evaluate the distribution
of accessibility across space and between social groups. Furthermore, in
the spirit of open data products
\citep{arribas-belOpenDataProductsA2021}, the package can be expanded
through collaboration with other researchers by, for example, including
travel time matrices to other essential destinations within the DMTI's
dataset (\emph{e.g.}, schools, healthcare, etc.). In other words, we
hope that by making these datasets publicly available, future analysis
can contribute to making Canada's transportation system more just and
fair, considering accessibility's as the main social good of
transportation \citep{martensTransportJusticeDesigning2016}, and the
inherent connection between public transit and the ``right to the city''
\citep{cogginRightTransportMoving2015}.
